An integrated distance sampling approach for modelling age-structured populations

### 2023-12-06 Manuscript_Nilsen_Nater.qmd
Updated text and bib-file

*Estimation of abundance and demographic rates for populations of wild vertebrate species is a challenging but fundamental issue for both management and research into ecology and evolution. One approach that has been used extensively to estimate abundance of wild living populations is Distance Sampling (DS) methods based on line transect survey data. Historically, DS models were only available as open population models, and did not allow for the direct estimation of changes in abundance through time. The advent of open population formulations based on the DS framework greatly extended the scope of the models, but so far models that estimate both temporal dynamics in abundance as well as the underlying demographic rates has not been implemented. Here, we present an integrated distance sampling apporach, that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the temporal variation in the demographic rates (recruitment rate and survival probability) that determine the temporal transition. The underlying process model is based on a two-stage transition matrix (*$A_t$*), with matrix entries given by survival probabilities (*$S$*) and time-specific recruitment rates (*$R_t$*). This framework allow us to make full use of the available data, and to effectively integrate the two data sources in an integrated modelling framework. Moreover, demographic rates often respond to environmental variation, and our approach allow us to directly estimate the effect of such environmental covariates on demographic rate variation. We first fit the model to simulated data, to assess it's ability to recover the underlying population dynamics when the true values are known. Then, we use data from a study of willow ptarmigan (Lagopus lagopus) in Norway as a case study.*

@@

Our open population integrated distance sampling model consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring (radio-telemetry in the example case).

@@

2.  Data from an individual based research project based on radio collared willow ptarmigan. The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2019 (and ongoing) and the individual based data spanning 2015-2019 (and ongoing).

Line transect survey data were collected in August each year, prior to the annual autumn harvest season. Line transects are surveyed by educated volunteers, using pointing dogs to located the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds as well as the age (juvenile and adult) and sex of the birds are recorded. \[Some more information on the surveys, e.g. why are individuals in groups, how are sexes and age classes distinguished, etc.\] Additional information about the surveys are found in Nilsen *et al.* (2020) and Bowler *et al.* (2020). \[Add sample sizes and additional information on data processing/storage post collection, i.e. how do volunteers report data, storage in hønsefuglportalen, quality control, pushes to LivingNorway and GBIF\].

The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2020. Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in Israelsen et al. (2020). To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds. After capture, birds were placed in an opaque bag to reduce stress. Before releasing the birds, they were fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g). The radio transmitters had an expected battery lifetime of 24 months (RI-2BM) or 30 months (RI-2DM), and included a mortality circuit that was activated if a bird had been immobile for 12 hours. We monitored the birds throughout the year, by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel. If a mortality signal was heard from the transmitter, we recovered it as soon as possible to determine cause of death. A number of birds dispersed out of the main study areas and was thus out of signal range for field personnel on the ground. To avoid loss of data, we conducted aerial triangulation using a helicopter or airplane three times a year (May, September and November) in the years 2016-2020.

@@

@@

![](Figures/SimCheck_Detects_Temps.png)

@@

Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set. As is evident from Figure X, the population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021. The increase was most distinct from 2016 and onward. Across years, mean survival probability for August - January ($S_1$) was estimated at `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$upper`), survival probability for February - July ($S_2$) was estimated as `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$upper`), and annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at `r filter(tidy_summary_stats, model_parameter =="Mu_S")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S")$upper`). Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021. Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability. While the mean (baseline) recruitment $\mu_R$ was estimated as `r filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_R")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_R")$upper`).

![**Figure X.** Estimated density of willow ptarmigans in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks 95% credible interval.](Figures/TimeSeries_popDens.png)

#### Vital rates

![**Figure X.** Posterior densities of A) seasonal survival, B) average annual survival, and C) recruitment rate. For the latter, the yellow distribution is for the intercept, representing a baseline recruitment rate when rodent occupancy is low. The turquoise distributions are for year-specific estimates or recruitment rate, with darker colors indicating later years.](Figures/PostDens_Mu_S_tR.png)

![**Figure X.** Estimated recruitment rate of willow ptarmigans in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks the 95% credible interval.](Figures/TimeSeries_rRep.png)

![**Figure X.** Predicted effect of rodent occupancy on recruitment rate. Solid line represent the posterior median, ribbon marks the 95% credible interval.](Figures/Rep_betaR.R.png)

#### Detection parameters

![**Figure X.** Predicted average detection probabilities for distance sampling surveys in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks the 95% credible interval.](Figures/TimeSeries_pDetect.png)

@@

>>>>

*Estimation of abundance and demographic rates for populations of wild vertebrate species is a challenging but fundamental issue for both management and research into ecology and evolution. One approach that has been used extensively to estimate abundance of wild living populations is Distance Sampling (DS) methods based on line transect survey data. Historically, DS models were only available as open population models, and did not allow for the direct estimation of changes in abundance through time. The advent of open population formulations based on the DS framework greatly extended the scope of the models, but so far models that estimate both temporal dynamics in abundance as well as the underlying demographic rates has not been implemented. Here, we present an integrated distance sampling approach, that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the temporal variation in the demographic rates (recruitment rate and survival probability) that determine the temporal transition. The underlying process model is based on a two-stage transition matrix (*$A_t$*), with matrix entries given by survival probabilities (*$S$*) and time-specific recruitment rates (*$R_t$*). This framework allow us to make full use of the available data, and to effectively integrate the two data sources in an integrated modelling framework. Moreover, demographic rates often respond to environmental variation, and our approach allow us to directly estimate the effect of such environmental covariates on demographic rate variation. We first fit the model to simulated data, to assess it's ability to recover the underlying population dynamics when the true values are known. Then, we use data from a study of willow ptarmigan (Lagopus lagopus) in Norway as a case study.*

@@

Our open population integrated distance sampling model consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring. In the example case, these auxiliary data come from a radio-telemetry study, but in principle other types of capture-recapture data can also be used.

@@

2.  Data from an individual based research project based on radio collared willow ptarmigan. The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2021 (and ongoing) and the individual based data spanning 2015-2021 (and ongoing).

Line transect survey data were collected in August each year, prior to the annual autumn harvest season, as part of the program "Hønsefuglportalen". Hønsefuglportalen is a national program for line transect surveys of tetraonid birds, and the effort is directed mainly to willow ptarmigan habitats. In our case study, we used data from the western part of Lierne municipality. Line transects are surveyed by educated volunteers, using pointing dogs to located the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds as well as the age (juvenile and adult) and sex of the birds are recorded. As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size. Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex. Since 2019, data has been collected through a mobile app tailormade for this project. Before 2019, field workers reported their data through a dedicated web portal. After data is collected, they go through several steps of quality control: first by local contacts and subsequently by personnel at NINA. The data are standardized based on the Darwin-Core standard (ref), and made publicly available as sampling-event data set published through GBIF [@Nilsen2023a]. For additional description of the data collection procedures, see [@Bowler2020]; [@Kvasnes2018][@Nilsen2023a].

The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2020. Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020]. To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds. After capture, birds were placed in an opaque bag to reduce stress. Before releasing the birds, they were fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g). The radio transmitters had an expected battery lifetime of 24 months (RI-2BM) or 30 months (RI-2DM), and included a mortality circuit that was activated if a bird had been immobile for 12 hours. We monitored the birds throughout the year, by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel. If a mortality signal was heard from the transmitter, we recovered it as soon as possible to determine cause of death. A number of birds dispersed out of the main study areas and was thus out of signal range for field personnel on the ground. To avoid loss of data, we conducted aerial triangulation using a helicopter or airplane three times a year (May, September and November) in the years 2016-2020.

@@

Based on simulated data, we assessed the models ability to estimate the underlying parameters. We run a total of 10 simulations, and the results are summarized in

@@

![Model performance on simulated data. Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability](Figures/SimCheck_Detects_Temps.png){#@fig-sim-detection}

@@

Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set. As is evident from Figure X, the population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021 (@fig-lierne-density). The increase was most distinct from 2016 and onward. Across years, mean survival probability for August - January ($S_1$) was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$upper, 2)`), survival probability for February - July ($S_2$) was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$upper, 2)`) (@fig-lierne-demo). Annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$upper, 2)`). Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021.

Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability. While the mean (baseline) recruitment $\mu_R$ was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean, 1)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$lower, 1)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$upper, 1)`). The estimated recruitment rate varied (as expected) among years: while the annual recruitment rate was estimated at `r round(max(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2007, it was estimated at `r round(min(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2012. Moreover, in contrast to the Alternative Prey Hypothesis (APH), our model did not find a clear link between the annual recruitment and the abundance of small rodents (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I. = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`).

![Estimated density of willow ptarmigans in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks 95% credible interval.](Figures/TimeSeries_popDens.png){#fig-lierne-density}

![Posterior densities of A) seasonal survival, B) average annual survival, and C) recruitment rate. For the latter, the yellow distribution is for the intercept, representing a baseline recruitment rate when rodent occupancy is low. The turquoise distributions are for year-specific estimates or recruitment rate, with darker colors indicating later years.](Figures/PostDens_Mu_S_tR.png){#fig-lierne-demo}

####

@@

-   In our process model, the population growth rate ($\lambda$) is determind by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, creating a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. As a test of the phenomena, we also
